Association between trauma severity models and opportunities for improvement:

### 2022-12-09 manuscript.Rmd
proofread method

The fitted curves for the logistic regression models of each score are shown in figure x. The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.

@@

The receiver operating curves for all scores are plotted in figure x. The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.

@@

The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively. The accuracy at all cutoffs is plotted for all scores in figure x. The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.

@@

The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths. Although this study was not specifically about preventable deaths, the factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e. whether there was a significant opportunity for improvement, which makes the previous studies on preventable deaths broadly relevant to this study. Ghorbani et al found that both TRISS and NORMIT had poor performance in predicting preventable deaths [@prevdeathsghorbani]. Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16]. Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI. The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.

Even though our results were expected, it was still important to investigate this question scientifically, because trauma severity models are in use today as audit filters at many institutions. Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful. Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed. Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity. They are much less useful when applied to individuals, and were not intended for that use.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pepe2019; @trentzsch2004], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].

>>>>

The fitted curves for the logistic regression models of each score are shown in figure 3. The odds ratios for TRISS, NORMIT and PS were `r t[['or']]` (95% CI: `r t[['or.ci.lo']]`-`r t[['or.ci.hi']]`, *p*: `r t[['or.p']]`), `r n[['or']]` (95% CI: `r n[['or.ci.lo']]`-`r n[['or.ci.hi']]`, *p*: `r n[['or.p']]`), and `r p[['or']]` (95% CI: `r p[['or.ci.lo']]`-`r p[['or.ci.hi']]`, *p*: `r p[['or.p']]`) respectively.

@@

The receiver operating curves for all scores are plotted in 4. The areas under the curve for TRISS, NORMIT and PS were `r t[['auc']]` (95% CI: `r t[['auc.ci.lo']]`-`r t[['auc.ci.hi']]`), `r n[['auc']]` (95% CI: `r n[['auc.ci.lo']]`-`r n[['auc.ci.hi']]`), and `r p[['auc']]` (95% CI: `r p[['auc.ci.lo']]`-`r p[['auc.ci.hi']]`) respectively.

@@

The integrated calibration index for TRISS, NORMIT and PS was `r t[['ici']]` (95% CI: `r t[['ici.ci']][1]`-`r t[['ici.ci']][2]`, `r n[['ici']]` (95% CI: `r n[['ici.ci']][1]`-`r n[['ici.ci']][2]` and `r p[['ici']]` (95% CI: `r p[['ici.ci']][1]`-`r p[['ici.ci']][2]` respectively. The accuracy at all cutoffs is plotted for all models in  5. The accuracy at a cutoff of 0.5 for TRISS, NORMIT and PS was `r t[['acc']]` (95% CI: `r t[['acc.ci']][1]`-`r t[['acc.ci']][2]`), `r n[['acc']]` (95% CI: `r n[['acc.ci']][1]`-`r n[['acc.ci']][2]`) and `r p[['acc']]` (95% CI: `r p[['acc.ci']][1]`-`r p[['acc.ci']][2]`) respectively.

@@

The results of this study agree with previous studies assessing the predictive power of trauma severity models with regards to preventable deaths. Although this study was not specifically about preventable deaths, the factor that determines whether a death was preventable is whether an improvement in care could have prevented the death, i.e. whether there was a significant opportunity for improvement, which makes the previous studies on preventable deaths broadly relevant to this study. Ghorbani et al found that both TRISS and NORMIT had poor performance in predicting preventable deaths [@prevdeathsghorbani]. Heim et al found that PS and TRISS miss a significant proportion of preventable deaths [@brohi16]. Our results also show that none of the three models investigated (TRISS, NORMIT and PS) were able to reliably predict OFI across the board and not only in mortalities. The difference between our study and prior studies is that we looked at the trauma population across the board and not only deaths.

Even though our results were expected, it was still important to investigate this topic scientifically because trauma severity models are in use today as audit filters. Our findings show that the use of trauma severity models as audit filters is unlikely to lead to improvements in case selection, and in fact could be harmful. Audit filters that don't improve case selection will either lead to cases being over-selected, causing resource wastage, or to cases being under-selected, leading to OFI being missed. Trauma severity models are useful when used on the population level, to allow institutions to compare themselves and to provide a framework for discussing trauma severity. They are much less useful when applied to individuals, and were not intended for that use.

@@

### Health equity
The demographical characteristics of the participants in this study reflect the well-known demographics of trauma patients worlwide. `r dd['male']`% of participants were male and the median age was `r dd['age']`, which matches with statistics both in Sweden and globally [@globalburden]. The predominance of male patients invites one to question if female patients are less likely to receive quality care. In healthcare as in all areas, repetition leads to mastery of a skill, so an onlooker might be tempted to assume that the high volume of male patients in trauma will lead healthcare providers to superiority iin treating males, whilst treatment of female patients will be neglected. This demographical difference has precipitated research that seeks to investigate whether this presumption holds true, and several large multi-centre studies have been published that evaluated whether there is a correlation between sex and outcomes. None of these studies found a correlation beween sex and increased mortality or other serious outcomes [@croce2002; @magnotti2008; @pape2019; @trentzsch2014], and in fact one study in the Netherlands found that female sex was correlated with decreased mortality [@pape2019].
